
```{r Package and data}

library(dplyr)
library(ggplot2)

knitr::opts_chunk$set(fig.width=6, fig.height=4, echo=FALSE, warning=FALSE, message=FALSE) 

lags <-  read.csv(file = "timeLags_all_variables_clean.csv",sep=";")
lags <- lags %>%
  filter(!is.na(timeLag) & off_FirstRec!=0) # we keep data only with both dates and with official first record later than 2000

# Change the name of some variables for clarityç

names(lags)[names(lags)=="ResearchEffort"] <- "NumAliensOff_Country"
names(lags)[names(lags)=="nOccTotSp"] <- "NumRecCS_Species"
names(lags)[names(lags)=="count"] <- "NumRecCS_SpCountry"
names(lags)[names(lags)=="obsEffort_LF"] <- "NumRecCS_CountryLifeform"
names(lags)[names(lags)=="mentioned_in_quarantine_species_regulations"] <- "quarantine"

```


```{r Preliminary data}
summary(lags)


barplot(table(lags$partly_native))
barplot(table(lags$Region),las=2)
barplot(table(lags$LifeForm),las=2)
barplot(table(lags$partly_native))
barplot(table(lags$habitat))
barplot(table(lags$eu_status))
barplot(table(lags$quarantine))
barplot(table(lags$worst_100_world_ISSG))
barplot(table(lags$worst_100_Europe_DAISIE))

# Dataset


ggplot(lags,aes(timeLag))+geom_histogram() + xlab("CSlags")


plot(lags$off_FirstRec,lags$firstRec)

# records per region
byregion <- lags %>%
  group_by(Region) %>%
  summarise(length(timeLag))
  
sd(byregion$`length(timeLag)`)/sqrt(length(byregion))

```

- habitat is very heterogeneous and they are mixed (one species can be marine and freshwater??). We group in two variables:
  (habitat2)can consider all with freshwater as aquatic. The species with a mixed of marine and other are from saline environment
  habitat3 separating aquatic from terrestrial
  
- Lifeform has many groups with extremely variable number of cases per level. 

- There are `r nrow(lags)` records with lag data. fully exotic `r  table(lags$partly_native)[1]`

- There is a long tail towards negative lags ( citizen science very early). Now they correspond to Belgium and Sweden. - Be aware of very early citizen science dates. Up to 1990

- A total of `r nrow(select(lags,timeLag>0)` 

Dataset
-  NumRecCS_Species (nOccTotSp) has been corrected and there aren't gaps
- Check Belgium data again (some 0 in official records)





```{r process data}
lags$Region <- as.factor(lags$Region)
lags$obsInNeigborCountryBefore <- as.factor(lags$obsInNeigborCountryBefore)

lags$habitat2 <- as.factor(lags$habitat)
summary(lags$habitat2)
levels(lags$habitat2) <- c("FW", "SAL","SAL","SAL","SAL","FW","MAR","SAL","SAL","TER","SAL","SAL","FW","SAL")
ggplot(lags,aes(y=timeLag,habitat2)) + geom_boxplot() 
barplot(table(lags$habitat2))
lags$habitat3 <- as.factor(lags$habitat)
levels(lags$habitat3) <- c("AQ", "AQ","AQ","AQ","AQ","AQ","AQ","AQ","AQ","TER","AQ","AQ","AQ","AQ")
ggplot(lags,aes(y=timeLag,habitat3)) + geom_boxplot() 

lags$LifeForm2 <- as.factor(lags$LifeForm)
ggplot(lags,aes(y=timeLag,LifeForm2)) + geom_boxplot() + theme(axis.text.x = element_text(angle = 90)) 

table(lags$LifeForm2)
levels(lags$LifeForm2) <- c("ALGAE","HERP","INV","MICRO","BIRD","INV","INV","FISH","MICRO","INS","INV","MAMMALS","INV","HERP","PLANTS")
ggplot(lags,aes(y=timeLag,LifeForm2)) + geom_boxplot() + theme(axis.text.x = element_text(angle = 90)) 

lags$LifeForm3 <- as.factor(lags$LifeForm)
levels(lags$LifeForm3) <- c("PLANTS&ALGAE","VERT","INV","MICRO","VERT","INV","INV","VERT","MICRO","INV","INV","VERT","INV","VERT","PLANTS&ALGAE")
ggplot(lags,aes(y=timeLag,LifeForm3)) + geom_boxplot() + theme(axis.text.x = element_text(angle = 90))

#create a variable indicating that at least ot is in one of the worst lists lists (ISSGS, DAISIE, EU)
lags$worst <- lags$worst_100_world_ISSG; table(lags$worst)
lags[lags$worst_100_Europe_DAISIE=="YES" &!is.na(lags$worst_100_Europe_DAISIE),"worst"] <- "YES" ; table(lags$worst)
lags[is.na(lags$worst),"worst"] <- "NO" # until we get more data we consider species without info not priority

lags$worst <- as.factor(lags$worst)


#create a variable indicating that at least the species is accepted or unders consideration in the UE regulation list
lags$eu_priority <- lags$eu_status; table(lags$eu_priority)
lags[lags$eu_status!="NO","eu_priority"] <- "YES" ; 

lags[is.na(lags$eu_priority),"eu_priority"] <- "NO" # until we get more data we consider species without info not priority

table(lags$eu_priority)
lags$eu_priority <- as.factor(lags$eu_priority)

table(lags$mentioned_in_quarantine_species_regulations)

# create variable for both worst and prioriy
lags$priority <- lags$eu_priority
lags[lags$worst=="YES" ,"priority"] <- "YES" ; table(lags$priority)

summary(lags)

lags$quarantine <- as.factor(lags$quarantine)


```

 - Create a new variable that indicates if the species is included in top list or EU concern/consideration. Now we have 32/668 species


```{r collinearity data}



variables <- c("habitat2","habitat3", "LifeForm2","LifeForm3","Region","priority","eu_priority","worst","quarantine","obsInNeigborCountryBefore", "google_sum","google_mean","google_country","scopus","NumRecCS_SpCountry","NumRecCS_CountryLifeform","NumAliensOff_Country","timeLag","NumRecCS_Species")

plot(lags[,variables[c(10:19)]])

for(i in 1:10) {
    name_var <- lags[,variables[i]]
    p<- ggplot(lags,aes(y=timeLag,x=(name_var))) + geom_boxplot() + theme(axis.text.x = element_text(angle = 90)) 
    print(p)
}

library(corrplot)

cormat <- cor(lags[,variables[11:19]])

corrplot(cormat,order = 'hclust',addrect = 3,addCoef.col = "black",tl.cex = 0.7,number.cex = 0.7)

library(lsr)
str(lags[,variables[1:10]])
cramersV(table(lags[,c("habitat2","LifeForm2")]))
cramersV(table(lags[,c("habitat3","LifeForm3")]))
cramersV(table(lags[,c("habitat2","LifeForm3")]))
cramersV(table(lags[,c("habitat3","LifeForm2")]))




# Scale variables to improve convergence
lags[,variables[c(11:17,19)]] <- scale(lags[,variables[c(11:17,19)]])




```


There isn't large correlation among continuous variables apart from google data
- Google_sum and google_mean are the same
- We can use the google_country or google_mean but not both together
- Least association between habitat and lifeform is using habitat2 and lifeform3




# Model with all variables

- N= 720 

```{r allvariables model}

# lme4
library(lme4)
library(multcomp)
library(lmerTest)
library(MuMIn);options(na.action = "na.fail")
library(knitr)
library(gridExtra)

modelbas <- lmer(timeLag~  google_country+ scopus + NumAliensOff_Country +habitat2+ LifeForm3 + obsInNeigborCountryBefore+NumRecCS_Species+NumRecCS_SpCountry+NumRecCS_CountryLifeform + eu_priority+worst+quarantine+(1|Region),data=lags )
summary(modelbas)

comp <- dredge(modelbas)
comp

result <- summary(model.avg(comp,subset = delta < 4))
MuMIn::sw(comp)
kable(result[10])
result
# str(result)

modelbest <- get.models(comp,subset = delta == 0)
summary(modelbest[[1]])
boxplot(lags$timeLag~lags$obsInNeigborCountryBefore)

# boxplot(lags$timeLag~lags$habitat2)
test <- glht(modelbest[[1]], linfct = mcp(habitat2 = "Tukey"))
summary(test) #no differences
groups_hab <- cld(summary(test));
p1 <- ggplot(lags,aes(x=(habitat2),y=timeLag)) + geom_violin() + xlab("Habitat") + ylab("CSlag")+
   geom_text(data=data.frame(), aes(x=levels(lags$habitat2),y=rep(max(lags$timeLag),4),label=groups_hab$mcletters$Letters), col='grey', size=8)

# boxplot(lags$timeLag~lags$LifeForm3)
test <- glht(modelbest[[1]], linfct = mcp(LifeForm3 = "Tukey"))
summary(test)  # significant differences between invertebrates and vertebrates
# groups <- cld(test)
groups_life <- cld(summary(test));
p2<- ggplot(lags,aes(x=(LifeForm3),y=timeLag)) + geom_violin() + xlab("Life form") + ylab("CSlag")+
   geom_text(data=data.frame(), aes(x=levels(lags$LifeForm3), y=rep(max(lags$timeLag),4),label=groups_life$mcletters$Letters), col='grey', size=8)

grid.arrange(p1,p2)

library(sjPlot)
library(sjlabelled)
library(sjmisc)
library(ggplot2)
set_theme(base = theme_light())
plot_model(modelbest[[1]],show.vales=T)
plot_model(modelbest[[1]],show.vales=T,terms = c("google_country","scopus","NumAliensOff_Country","obsInNeigborCountryBefore","NumRecCS_Species","NumRecCS_CountryLifeform","eu_priority","worst","quarantine") )

plot_model(modelbest[[1]],type="re",show.vales=T,title = " ",sort.est = "sort.all",grid = F)



```

- Citizens are earlier when it has been observed in the country before (more negative), It is the most important
- Faster for vertebrates than plants
- Faster when there are more citizen records of the species
- No effect of google or scopus hits
- being in a priority list is significant. 


- obs in neighbouring country is significant
- more records in scopus, faster is CS
- in vertebrates CS is faster



# Interaction plots

```{r interaction plots}
library(plotly)
p <- ggplot(data=lags,aes(y=timeLag,x=NumRecCS_Species,colour=LifeForm3)) + geom_point() + theme(axis.text.x = element_text(angle = 90)) + facet_grid(rows=vars(eu_priority),cols=vars(obsInNeigborCountryBefore),labeller = label_both)
p
# ggplotly(p)


ggplot(data=lags,aes(y=timeLag,x=NumRecCS_Species,colour=LifeForm3)) + geom_point() + theme(axis.text.x = element_text(angle = 90)) + facet_grid(rows=vars(eu_priority),cols=vars(quarantine),labeller = label_both)

ggplot(data=lags,aes(y=timeLag,x=NumRecCS_Species,colour=LifeForm3)) + geom_point() + theme(axis.text.x = element_text(angle = 90)) + facet_grid(rows=vars(worst),cols=vars(quarantine),labeller = label_both)

```
- Row: eu_species ; column: observed in Neigbouring country
- It seems the effect of eu_species of concern is mostly due to Cydalima perspectalis (under consideration)




# Model with only exotic to EU

- N= 504 

```{r EU model}

lags_eu <- lags %>%
  filter(partly_native=="NO" & Region!="New Zealand" )

summary(lags_eu)

# lme4
library(lme4)
library(multcomp)
library(lmerTest)
library(MuMIn);options(na.action = "na.fail")

modelbas <- lmer(timeLag~  google_country+ scopus + NumAliensOff_Country +habitat2+ LifeForm3 + obsInNeigborCountryBefore+NumRecCS_Species+NumRecCS_CountryLifeform + eu_priority+worst+quarantine+(1|Region),data=lags_eu )
summary(modelbas)

comp <- dredge(modelbas)
comp

summary(model.avg(comp,subset = delta < 4))

boxplot(lags_eu$timeLag~lags_eu$habitat2)
test <- glht(modelbas, linfct = mcp(habitat2 = "Tukey"))
summary(test) #no differences

boxplot(lags_eu$timeLag~lags_eu$LifeForm3)
test <- glht(modelbas, linfct = mcp(LifeForm3 = "Tukey"))
summary(test)  # significant differences between invertebrates and vertebrates
# groups <- cld(test)

library(sjPlot)
plot_model(modelbas,show.vales=T)
plot_model(modelbas,show.vales=T,type="re")



```

- obs in neighbouring country is significant
- more number of aliens species in official country, faster CS
- CS is faster for vertebrates than invertebrates or plants


# Model with only Belgium

- N= 353  
- no species in the quarantine list
```{r Belgium model}

lags_bel <- lags %>%
  filter(Region=="Belgium" ) %>%
  droplevels()

summary(lags_bel)
table(lags_bel$habitat2) # almost all terrestrial. not consider habitat
table(lags_bel$LifeForm3) # almost all plants
# lme4
library(lme4)
library(multcomp)
library(lmerTest)
library(MuMIn);options(na.action = "na.fail")

modelbas <- lm(timeLag~  google_country+ scopus + habitat2 + LifeForm3 + obsInNeigborCountryBefore+NumRecCS_Species + eu_priority+worst,data=lags_bel )
summary(modelbas)

comp <- dredge(modelbas)
comp

summary(model.avg(comp,subset = delta < 4))

boxplot(lags_bel$timeLag~lags_bel$habitat2)
library(emmeans)
pairs(emmeans(modelbas, "habitat2"))

boxplot(lags_bel$timeLag~lags_bel$LifeForm3)
pairs(emmeans(modelbas, "LifeForm3"))


boxplot(lags_bel$timeLag~lags_bel$obsInNeigborCountryBefore)

library(sjPlot)
plot_model(modelbas,show.vales=T)


```

- There are very few records per lifeform. Mostly all plants. So comparison might not be relevant. Anyway, vertebrates are
faster than invertebrates or plants 
- Faster when observed in neighbour country and when there are more CS records

# Model without Belgium

- N= 379 

```{r No Belgium model}

lags_nobel <- lags %>%
  filter(Region!="Belgium") %>%
  droplevels()

summary(lags_nobel)
table(lags_nobel$habitat2) # almost all terrestrial. not consider habitat
table(lags_nobel$LifeForm3) # almost all plants
# lme4
library(lme4)
library(multcomp)
library(lmerTest)
library(MuMIn);options(na.action = "na.fail")

modelbas <- lmer(timeLag~  google_country+ scopus + NumAliensOff_Country +habitat2+ LifeForm3 + obsInNeigborCountryBefore+NumRecCS_Species+NumRecCS_CountryLifeform + eu_priority+worst+qurantine+(1|Region),data=lags_nobel )
summary(modelbas)

comp <- dredge(modelbas)
comp

summary(model.avg(comp,subset = delta < 4))

boxplot(lags_nobel$timeLag~lags_nobel$habitat2)

boxplot(lags_nobel$timeLag~lags_nobel$eu_priority)
table(lags_nobel$eu_priority,lags_nobel$LifeForm3)
table(lags_nobel$eu_status,lags_nobel$LifeForm3)

library(emmeans)
pairs(emmeans(modelbas, "habitat2"))

boxplot(lags_nobel$timeLag~lags_nobel$LifeForm3)
pairs(emmeans(modelbas, "LifeForm3"))


boxplot(lags_nobel$timeLag~lags_nobel$obsInNeigborCountryBefore)

library(sjPlot)
plot_model(modelbas,show.vales=T)
plot_model(modelbas,show.vales=T,type="re")


```

# Model with balance number of records per country

- Still we need to get randomized results. Just one iteration at the moment

```{r Balanced model, eval=F}

mean(table(lags$Region[lags$Region!="Belgium"]))
# 12 records per country without Belgium
lags_bal <- lags %>%
  group_by(Region) %>%
  slice_sample(n = 12)



summary(lags_bal)
table(lags_bal$habitat2) # almost all terrestrial. not consider habitat
table(lags_bal$LifeForm3) # almost all plants
# lme4
library(lme4)
library(multcomp)
library(lmerTest)
library(MuMIn);options(na.action = "na.fail")

modelbas <- lmer(timeLag~  google_country+ scopus + habitat2 + LifeForm3 + obsInNeigborCountryBefore+NumRecCS_Species + priority + (1|Region),data=lags_bal )
summary(modelbas)

comp <- dredge(modelbas)
comp

summary(model.avg(comp,subset = delta < 4))

boxplot(lags_bal$timeLag~lags_bal$habitat2)
library(emmeans)
pairs(emmeans(modelbas, "habitat2"))

boxplot(lags_bal$timeLag~lags_bal$LifeForm3)
pairs(emmeans(modelbas, "LifeForm3"))


boxplot(lags_bel$timeLag~lags_bel$obsInNeigborCountryBefore)

library(sjPlot)
plot_model(modelbas,show.vales=T)
plot_model(modelbas,show.vales=T,type="re")

```

# Model only with data with positive lags
```{r positive model}

lags2 <- lags %>%
  filter(timeLag>0)

hist(lags2$timeLag)

modelpos <- glmer(timeLag~  google_country+ scopus + NumAliensOff_Country +habitat2+ LifeForm3 + obsInNeigborCountryBefore+NumRecCS_Species+NumRecCS_CountryLifeform + eu_priority+worst+ (1|Region),data=lags2,family="poisson" )
summary(modelpos)

#################Probar overdispersion - https://bbolker.github.io/mixedmodels-misc/glmmFAQ.html#fitting-models-with-overdispersion

overdisp_fun <- function(model) {
    rdf <- df.residual(model)
    rp <- residuals(model,type="pearson")
    Pearson.chisq <- sum(rp^2)
    prat <- Pearson.chisq/rdf
    pval <- pchisq(Pearson.chisq, df=rdf, lower.tail=FALSE)
    c(chisq=Pearson.chisq,ratio=prat,rdf=rdf,p=pval)
}

overdisp_fun(modelpos) # high and significant

## extract summary table; you may also be able to do this via
##  broom::tidy or broom.mixed::tidy
quasi_table <- function(model,ctab=coef(summary(model)),
                           phi=overdisp_fun(model)["ratio"]) {
    qctab <- within(as.data.frame(ctab),
    {   `Std. Error` <- `Std. Error`*sqrt(phi)
        `z value` <- Estimate/`Std. Error`
        `Pr(>|z|)` <- 2*pnorm(abs(`z value`), lower.tail=FALSE)
    })
    return(qctab)
}
printCoefmat(quasi_table(modelpos),digits=3)

# comp <- dredge(modelpos)
# comp
# 
# summary(model.avg(comp,subset = delta < 4))

boxplot(lags2$timeLag~lags2$habitat2)
test <- glht(modelpos, linfct = mcp(habitat2 = "Tukey"))
summary(test) #no differences

boxplot(lags2$timeLag~lags2$LifeForm3)
test <- glht(modelpos, linfct = mcp(LifeForm3 = "Tukey"))
summary(test)  # significant differences between invertebrates and vertebrates
# groups <- cld(test)

library(sjPlot)
plot_model(modelpos,show.vales=T)
plot_model(modelpos,show.vales=T,type="re")



```


- Think about the rationality 
- Zero-inflated and it is truncated. maybe we should use a truncated regression
-The key effect is from neighbouring country. 
- ACcounting for overdispersion ther isn't any other significant effect. 

Next steps
- maybe logs of scopus/ google variables
- mosaic plot between positive / negative / records
- update belgium data lags
- Randomize and select similar number of records per country. 


# Model only with data with positive lags for belgium
```{r positive model belgium}

lags3 <- lags %>%
  filter(timeLag>0 & Region!="Belgium")

hist(lags3$timeLag)

modelpos <- glmer(timeLag~  google_country+ scopus + NumAliensOff_Country +habitat2+ LifeForm3 + obsInNeigborCountryBefore+NumRecCS_Species+NumRecCS_CountryLifeform + eu_priority+worst+ (1|Region),data=lags3,family="poisson" )
summary(modelpos)

#################Probar overdispersion - https://bbolker.github.io/mixedmodels-misc/glmmFAQ.html#fitting-models-with-overdispersion

overdisp_fun <- function(model) {
    rdf <- df.residual(model)
    rp <- residuals(model,type="pearson")
    Pearson.chisq <- sum(rp^2)
    prat <- Pearson.chisq/rdf
    pval <- pchisq(Pearson.chisq, df=rdf, lower.tail=FALSE)
    c(chisq=Pearson.chisq,ratio=prat,rdf=rdf,p=pval)
}

overdisp_fun(modelpos) # high and significant

## extract summary table; you may also be able to do this via
##  broom::tidy or broom.mixed::tidy
quasi_table <- function(model,ctab=coef(summary(model)),
                           phi=overdisp_fun(model)["ratio"]) {
    qctab <- within(as.data.frame(ctab),
    {   `Std. Error` <- `Std. Error`*sqrt(phi)
        `z value` <- Estimate/`Std. Error`
        `Pr(>|z|)` <- 2*pnorm(abs(`z value`), lower.tail=FALSE)
    })
    return(qctab)
}
printCoefmat(quasi_table(modelpos),digits=3)

# comp <- dredge(modelpos)
# comp
# 
# summary(model.avg(comp,subset = delta < 4))

boxplot(lags3$timeLag~lags3$habitat2)
test <- glht(modelpos, linfct = mcp(habitat2 = "Tukey"))
summary(test) #no differences

boxplot(lags3$timeLag~lags3$LifeForm3)
test <- glht(modelpos, linfct = mcp(LifeForm3 = "Tukey"))
summary(test)  # significant differences between invertebrates and vertebrates
# groups <- cld(test)

library(sjPlot)
plot_model(modelpos,show.vales=T)
plot_model(modelpos,show.vales=T,type="re")



```


# Model truncated years

 <=2015 551 observations

```{r truncated}

lags_trunc <- lags %>%
  filter(off_FirstRec<=2015) %>%
  droplevels()

summary(lags_trunc)
table(lags_nobel$habitat2) # almost all terrestrial. not consider habitat
table(lags_nobel$LifeForm3) # almost all plants


modelbas <- lmer(timeLag~  google_country+ scopus + NumAliensOff_Country +habitat2+ LifeForm3 + obsInNeigborCountryBefore+NumRecCS_Species+NumRecCS_CountryLifeform + eu_priority+worst+quarantine+(1|Region),data=lags_trunc )
summary(modelbas)

comp <- dredge(modelbas)
comp

summary(model.avg(comp,subset = delta < 4))

boxplot(lags_trunc$timeLag~lags_trunc$habitat2)

boxplot(lags_trunc$timeLag~lags_trunc$eu_priority)
table(lags_trunc$eu_priority,lags_trunc$LifeForm3)
table(lags_trunc$eu_status,lags_trunc$LifeForm3)

pairs(emmeans(modelbas, "habitat2"))

boxplot(lags_trunc$timeLag~lags_trunc$LifeForm3)
pairs(emmeans(modelbas, "LifeForm3"))


boxplot(lags_trunc$timeLag~lags_trunc$obsInNeigborCountryBefore)


plot_model(modelbas,show.vales=T)
plot_model(modelbas,show.vales=T,type="re")


```