
```{r Package and data}

library(dplyr)
library(ggplot2)


lags <-  read.csv(file = "timeLags_all_variables_clean.csv",sep=";")
lags <- lags %>%
  filter(!is.na(timeLag)) # we keep data only with both dates and with official first record later than 2000

# Change the name of some variables for clarity

names(lags)[names(lags)=="ResearchEffort"] <- "NumAliensOff_Country"
names(lags)[names(lags)=="nOccTotSp"] <- "NumRecCS_Species"
names(lags)[names(lags)=="obsEffort_LF"] <- "NumRecCS_CountryLifeform"

```


```{r Preliminary data}
summary(lags)


table(lags$partly_native)
table(lags$Region)
table(lags$LifeForm)
table(lags$partly_native)
table(lags$habitat)
table(lags$eu_status)
table(lags$mentioned_in_EU_quarantine_species)
table(lags$worst_100_world_ISSG)
table(lags$worst_100_Europe_DAISIE)
table(lags$mentioned_in_EU_quarantine_species)
# Dataset

hist(lags$timeLag)

plot(lags$off_FirstRec,lags$firstRec)




```

- habitat is very heterogeneous and they are mixed (one species can be marine and freshwater??). We group in two variables:
  (habitat2)can consider all with freshwater as aquatic. The species with a mixed of marine and other are from saline environment
  habitat3 separating aquatic from terrestrial
  
- Lifeform has many groups with extremely variable number of cases per level. 

- There are `r nrow(lags)` records with lag data. fully exotic `r  table(lags$partly_native)[1]`

- There is a long tail towards negative lags ( citizen science very early). Now they correspond to Belgium and Sweden. - Be aware of very early citizen science dates. Up to 1990

Dataset
- Check NumRecCS_Species (nOccTotSp) for NA values. There are 155 records without this variable. Not using it at the moment
- There are some species with NA in top lists. Double check species_variable.csv

- separating in two models depending on the tail
- belgium overrepresented. Check with Tim. 



```{r process data}
lags$Region <- as.factor(lags$Region)
lags$obsInNeigborCountryBefore <- as.factor(lags$obsInNeigborCountryBefore)

lags$habitat2 <- as.factor(lags$habitat)
levels(lags$habitat2) <- c("FW", "SAL","SAL","SAL","FW","MAR","TER","SAL","SAL","FW")
ggplot(lags,aes(y=timeLag,habitat2)) + geom_boxplot() 

lags$habitat3 <- as.factor(lags$habitat)
levels(lags$habitat3) <- c("AQ", "AQ","AQ","AQ","AQ","AQ","TER","AQ","AQ","AQ")
ggplot(lags,aes(y=timeLag,habitat3)) + geom_boxplot() 

lags$LifeForm2 <- as.factor(lags$LifeForm)
ggplot(lags,aes(y=timeLag,LifeForm2)) + geom_boxplot() + theme(axis.text.x = element_text(angle = 90)) 

table(lags$LifeForm2)
levels(lags$LifeForm2) <- c("ALGAE","HERP","INV","MICRO","BIRD","INV","INV","FISH","MICRO","INS","INV","MAMMALS","INV","HERP","PLANTS")
ggplot(lags,aes(y=timeLag,LifeForm2)) + geom_boxplot() + theme(axis.text.x = element_text(angle = 90)) 

lags$LifeForm3 <- as.factor(lags$LifeForm)
levels(lags$LifeForm3) <- c("PLANTS&ALGAE","VERT","INV","MICRO","VERT","INV","INV","VERT","MICRO","INV","INV","VERT","INV","VERT","PLANTS&ALGAE")
ggplot(lags,aes(y=timeLag,LifeForm3)) + geom_boxplot() + theme(axis.text.x = element_text(angle = 90))

#create a variable indicating that at least ot is in one of the priority lists (ISSGS, DAISIE, EU)
lags$priority <- lags$worst_100_world_ISSG; table(lags$priority)
lags[lags$worst_100_Europe_DAISIE=="YES" &!is.na(lags$worst_100_Europe_DAISIE),"priority"] <- "YES" ; table(lags$priority)
lags[lags$eu_status!="NO","priority"] <- "YES" ; 

lags[is.na(lags$priority),"priority"] <- "NO" # until we get more data we consider species without info not priority

table(lags$priority)
lags$priority <- as.factor(lags$priority)




summary(lags)

```

 - Create a new variable that indicates if the species is included in top list or EU concern/consideration. Now we have 32/668 species


```{r collinearity data}



variables <- c("habitat2","habitat3", "LifeForm2","LifeForm3","Region","priority","obsInNeigborCountryBefore", "google_sum","google_mean","google_country","scopus","NumRecCS_CountryLifeform","NumAliensOff_Country","timeLag","NumRecCS_Species")

plot(lags[,variables[8:14]])

for(i in 1:7) {
    name_var <- lags[,variables[i]]
    p<- ggplot(lags,aes(y=timeLag,x=(name_var))) + geom_boxplot() + theme(axis.text.x = element_text(angle = 90)) 
    print(p)
}

library(corrplot)

cormat <- cor(lags[,variables[8:14]])

corrplot(cormat,order = 'hclust',addrect = 3,addCoef.col = "black")

library(lsr)
str(lags[,variables[1:7]])
cramersV(table(lags[,c("habitat2","LifeForm2")]))
cramersV(table(lags[,c("habitat3","LifeForm3")]))
cramersV(table(lags[,c("habitat2","LifeForm3")]))
cramersV(table(lags[,c("habitat3","LifeForm2")]))




# Scale variables to improve convergence
lags[,variables[c(8:13,15)]] <- scale(lags[,variables[c(8:13,15)]])



# considering also NumRecCS

lags_full <- filter(lags,!is.na(NumRecCS_Species))

cormat <- cor(lags_full[,variables[8:15]])

corrplot(cormat,order = 'hclust',addrect = 3,addCoef.col = "black")

```


There isn't large correlation among continuous variables apart from google data
- Google_sum and google_mean are the same
- We can use the google_country or google_mean but not both together

- Least association between habitat and lifeform is using habitat2 and lifeform3

- When considering all no empty records NumRecCS shows strong effect. Worth filling the gaps!



# Model with all variables

- N= 568. as we have gaps in NumRecCS_Species

```{r allvariables model}

# lme4
library(lme4)
library(multcomp)
library(lmerTest)
library(MuMIn);options(na.action = "na.fail")

modelbas <- lmer(timeLag~  google_country+ scopus + NumAliensOff_Country +habitat2+ LifeForm3 + obsInNeigborCountryBefore+NumRecCS_Species+NumRecCS_CountryLifeform + priority+(1|Region),data=lags_full )
summary(modelbas)

comp <- dredge(modelbas)
comp

summary(model.avg(comp,subset = delta < 4))

boxplot(lags_full$timeLag~lags_full$habitat2)
test <- glht(modelbas, linfct = mcp(habitat2 = "Tukey"))
summary(test) #no differences

boxplot(lags_full$timeLag~lags_full$LifeForm3)
test <- glht(modelbas, linfct = mcp(LifeForm3 = "Tukey"))
summary(test)  # significant differences between invertebrates and vertebrates
# groups <- cld(test)

library(sjPlot)
plot_model(modelbas,show.vales=T)
plot_model(modelbas,show.vales=T,type="re")



```

- Citizens are earlier when it has been observed in the country before (more negative), It is the most important
- Faster for vertebrates than plants
- Faster when there are more citizen records of the species
- No effect of google or scopus hits
- being in a priority list is marginally significant. worth completing gaps

# Model with all data

- N= 723 not considering NumRecCS_Species as it has >100

```{r basic model}

# lme4
library(lme4)
library(multcomp)
library(lmerTest)
library(MuMIn);options(na.action = "na.fail")

modelbas <- lmer(timeLag~  google_country+ scopus + NumAliensOff_Country +habitat2+ LifeForm3 + obsInNeigborCountryBefore+NumRecCS_CountryLifeform + priority+(1|Region),data=lags )
summary(modelbas)

comp <- dredge(modelbas)
comp

summary(model.avg(comp,subset = delta < 4))

boxplot(lags$timeLag~lags$habitat2)
test <- glht(modelbas, linfct = mcp(habitat2 = "Tukey"))
summary(test) #no differences

boxplot(lags$timeLag~lags$LifeForm3)
test <- glht(modelbas, linfct = mcp(LifeForm3 = "Tukey"))
summary(test)  # significant differences between invertebrates and vertebrates
# groups <- cld(test)

library(sjPlot)
plot_model(modelbas,show.vales=T)
plot_model(modelbas,show.vales=T,type="re")



```

- obs in neighbouring country is significant
- more records in scopus, faster is CS
- in vertebrates CS is faster



# Model with only exotic to EU

- N= 500 not considering NumRecCS_Species as it has >100

```{r EU model}

lags_eu <- lags %>%
  filter(partly_native=="NO" & Region!="New Zealand" )

summary(lags_eu)

# lme4
library(lme4)
library(multcomp)
library(lmerTest)
library(MuMIn);options(na.action = "na.fail")

modelbas <- lmer(timeLag~  google_country+ scopus + NumAliensOff_Country +habitat2+ LifeForm3 + obsInNeigborCountryBefore+NumRecCS_CountryLifeform + priority+(1|Region),data=lags_eu )
summary(modelbas)

comp <- dredge(modelbas)
comp

summary(model.avg(comp,subset = delta < 4))

boxplot(lags_eu$timeLag~lags_eu$habitat2)
test <- glht(modelbas, linfct = mcp(habitat2 = "Tukey"))
summary(test) #no differences

boxplot(lags_eu$timeLag~lags_eu$LifeForm3)
test <- glht(modelbas, linfct = mcp(LifeForm3 = "Tukey"))
summary(test)  # significant differences between invertebrates and vertebrates
# groups <- cld(test)

library(sjPlot)
plot_model(modelbas,show.vales=T)
plot_model(modelbas,show.vales=T,type="re")



```

- obs in neighbouring country is significant
- more number of aliens species in official country, faster CS
- CS is faster for vertebrates than invertebrates or plants


# Model with only Belgium

- N= 319  considering NumRecCS_Species as it has only 25 empty species

```{r Belgium model}

lags_bel <- lags %>%
  filter(Region=="Belgium" & !is.na(NumRecCS_Species)) %>%
  droplevels()

summary(lags_bel)
table(lags_bel$habitat2) # almost all terrestrial. not consider habitat
table(lags_bel$LifeForm3) # almost all plants
# lme4
library(lme4)
library(multcomp)
library(lmerTest)
library(MuMIn);options(na.action = "na.fail")

modelbas <- lm(timeLag~  google_country+ scopus + habitat2 + LifeForm3 + obsInNeigborCountryBefore+NumRecCS_Species + priority,data=lags_bel )
summary(modelbas)

comp <- dredge(modelbas)
comp

summary(model.avg(comp,subset = delta < 4))

boxplot(lags_bel$timeLag~lags_bel$habitat2)
library(emmeans)
pairs(emmeans(modelbas, "habitat2"))

boxplot(lags_bel$timeLag~lags_bel$LifeForm3)
pairs(emmeans(modelbas, "LifeForm3"))


boxplot(lags_bel$timeLag~lags_bel$obsInNeigborCountryBefore)

library(sjPlot)
plot_model(modelbas,show.vales=T)


```

- There are very few records per lifeform. Mostly all plants. So comparison might not be relevant. Anyway, vertebrates are
faster than invertebrates or plants 
- Faster when observed in neighbour country and when there are more CS records


# Model with balance number of records per country

- Still we need to get randomized results. Just one iteration at the moment

```{r Balanced model, eval=F}

mean(table(lags$Region[lags$Region!="Belgium"]))
# 12 records per country without Belgium
lags_bal <- lags %>%
  group_by(Region) %>%
  slice_sample(n = 12)



summary(lags_bal)
table(lags_bal$habitat2) # almost all terrestrial. not consider habitat
table(lags_bal$LifeForm3) # almost all plants
# lme4
library(lme4)
library(multcomp)
library(lmerTest)
library(MuMIn);options(na.action = "na.fail")

modelbas <- lmer(timeLag~  google_country+ scopus + habitat2 + LifeForm3 + obsInNeigborCountryBefore+NumRecCS_Species + priority + (1|Region),data=lags_bal )
summary(modelbas)

comp <- dredge(modelbas)
comp

summary(model.avg(comp,subset = delta < 4))

boxplot(lags_bal$timeLag~lags_bal$habitat2)
library(emmeans)
pairs(emmeans(modelbas, "habitat2"))

boxplot(lags_bal$timeLag~lags_bal$LifeForm3)
pairs(emmeans(modelbas, "LifeForm3"))


boxplot(lags_bel$timeLag~lags_bel$obsInNeigborCountryBefore)

library(sjPlot)
plot_model(modelbas,show.vales=T)
plot_model(modelbas,show.vales=T,type="re")

```

# Model only with data with positive lags
```{r positive model}

lags2 <- lags %>%
  filter(timeLag>=0)

hist(lags2$timeLag)

modelpos <- glmer(timeLag~  google_country+ scopus + NumAliensOff_Country +habitat2+ LifeForm3 + obsInNeigborCountryBefore+NumRecCS_CountryLifeform + priority+ (1|Region),data=lags2,family="poisson" )
summary(modelpos)

#################Probar overdispersion - https://bbolker.github.io/mixedmodels-misc/glmmFAQ.html#fitting-models-with-overdispersion

overdisp_fun <- function(model) {
    rdf <- df.residual(model)
    rp <- residuals(model,type="pearson")
    Pearson.chisq <- sum(rp^2)
    prat <- Pearson.chisq/rdf
    pval <- pchisq(Pearson.chisq, df=rdf, lower.tail=FALSE)
    c(chisq=Pearson.chisq,ratio=prat,rdf=rdf,p=pval)
}

overdisp_fun(modelpos) # high and significant

## extract summary table; you may also be able to do this via
##  broom::tidy or broom.mixed::tidy
quasi_table <- function(model,ctab=coef(summary(model)),
                           phi=overdisp_fun(model)["ratio"]) {
    qctab <- within(as.data.frame(ctab),
    {   `Std. Error` <- `Std. Error`*sqrt(phi)
        `z value` <- Estimate/`Std. Error`
        `Pr(>|z|)` <- 2*pnorm(abs(`z value`), lower.tail=FALSE)
    })
    return(qctab)
}
printCoefmat(quasi_table(modelpos),digits=3)

# comp <- dredge(modelpos)
# comp
# 
# summary(model.avg(comp,subset = delta < 4))

boxplot(lags2$timeLag~lags2$habitat2)
test <- glht(modelpos, linfct = mcp(habitat2 = "Tukey"))
summary(test) #no differences

boxplot(lags2$timeLag~lags2$LifeForm3)
test <- glht(modelpos, linfct = mcp(LifeForm3 = "Tukey"))
summary(test)  # significant differences between invertebrates and vertebrates
# groups <- cld(test)

library(sjPlot)
plot_model(modelpos,show.vales=T)
plot_model(modelpos,show.vales=T,type="re")



```


- Think about the rationality 
- Zero-inflated and it is truncated. maybe we should use a truncated regression
-The key effect is from neighbouring country. 
- ACcounting for overdispersion ther isn't any other significant effect. 

Next steps
- Check gaps in data
- Randomize and select similar number of records per country. 


